Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-09-19 prop_letter.tex
prop_letter.tex Add comments...

In the remainder of this section, we will show that our definition of
QIVPMs is consistent with classical IVPMs in the sense that a
restriction of QIVPMs to commuting events recovers the definition of
classical IVPMs. To formalize this connection between QIVPMs and
classical IVPMs, we first note that in contrast to the case with
conventional quantum probability spaces, there is generally no unique
state~$\rho$ consistent with a QIVPM~\citep{HOSTunpublished}. We say that a QIVPM $\bar{\mu}$
is consistent with a state $\rho$ and projector~$P$ if the interval
$\bar{\mu}(P)$ contains the exact probability calculated by the Born
rule, i.e, if:
\begin{equation}
\mu_{\rho}^{\mathrm{B}}\left(P\right)\in\bar{\mu}\left(P\right)
\end{equation}
where $\mu_{\rho}^{\mathrm{B}}\left(P\right)=\Tr\left(\rho P\right)$
is the probability that the event~$P$ is determined by the Born
rule~\citep{Born1983,peres1995quantum,544199}.

@@

This theorem follows directly from our
correspondence~\cite{HOSTunpublished} and Shapley's
theorem~\cite{Shapley1971,GilboaSchmeidler1994,Grabisch2016}, which
proves that a classical IVPM always contains at least one state $\rho$
that is consistent with every event in $\events_C$. In contrast, as we
illustrate in the next section, the set of states
$\coreBorn\left(\bar{\mu},\events'\right)$ that is consistent with
some arbitrary (not necessarily commuting) events $\events'$ and a
particular QIVPM $\bar{\mu}$ may be \emph{empty}.

@@

\noindent Intuitively the expectation value of an observable relative
to a QIVPM $\bar{\mu}$ lies between two possible outcomes, which
themselves lie between the minimum and maximum bounds of the probability
intervals associated with each state~$\rho$ that is consistent with
$\bar{\mu}$ on every projector in the spectral decomposition of the
observable. If $\bar{\mu}$ is a conventional (Born) probability measure
induced by a state $\rho$, then the Born rule probability induced
by every state in $\coreBorn\left(\bar{\mu},\events'\right)$ will
be $\mu_{\rho}^{\mathrm{B}}$ and the interval collapses to a point,
thus reducing the definition to that of Eq.~(\ref{eq:quantum-expectation}).
For example, we can extend one of the fragments of valid probability
measures~$\mu_{1}$ defined in Table~\ref{tab:quantum-probability-measure}
to a QIVPM~$\bar{\mu}_{3}$:
\begin{equation}
\begin{aligned}\bar{\mu}_{3}(\proj{0}) & =\left[\tfrac{1}{2},\tfrac{1}{2}\right] & \bar{\mu}_{3}(\proj{2}) & =\left[0,0\right]\\
\bar{\mu}_{3}(\proj{1}) & =\left[\tfrac{1}{2},\tfrac{1}{2}\right] &  & \ldots\textrm{ etc.}
\end{aligned}
\label{eq:Born-QIVPM}
\end{equation}
For the same
observable~$\mathbf{O}=\proj{0}+2\,\proj{1}+3\,\proj{2}$, the only
state~$\rho$ consistent with $\bar{\mu}_{3}$ on $\proj{0}$, $\proj{1}$,
and $\proj{2}$ must be $\proj{\ps}$, and $\mu_{1}=\mu_{\proj{\ps}}^{\mathrm{B}}$.
Therefore, the expectation value~$\expval{\mathbf{O}}_{\bar{\mu}_{3}}=\left[\expval{\mathbf{O}}_{\mu_{1}},\expval{\mathbf{O}}_{\mu_{1}}\right]=\left[1.5,1.5\right]$
which is an interval, but not between $0$ and $1$ because it is
not a probability of an event. We also note that, when restricted
to commuting projectors, the definition above is consistent with the
classical notion of the \emph{Choquet integral}~\citep{GilboaSchmeidler1994,Grabisch2016}
which is used to calculate the expectation value of random variables
as a weighted average.

@@

\noindent The definition puts no restrictions on the choice of intervals,
only on the assignment of intervals to events. When $\delta=0$, every
event must be assigned a probability either in the interval $\imposs$
or in $\necess$, i.e., every event is completely determined with
certainty. As $\delta$ gets larger, the QIVPM allows for more indeterminate
behavior. To illustrate the example easier, we restrict ourselves
to the classical subspace of commuting events~$\events_{0}=\left\{ \mathbb{0},\proj{0},\proj{1},\mathbb{1}\right\} $.
If we apply the same definition in the classical case, the classical
IVPM~$\bar{\mu}_{4}$ defined in Table~\ref{tab:classical-IVPMs}
is $0$-deterministic because $\imposs\subseteq\left[0,0\right]$
and $\necess\subseteq\left[1-0,1\right]$.
\begin{table}
\noindent \centering{}\caption{\label{tab:classical-IVPMs}Classical IVPMs~$\bar{\mu}_{4}$, $\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ on $\events_{0}$.}
\begin{tabular}{cccc}
\toprule 
\addlinespace
$P$ & $\bar{\mu}_{4}(P)$ & $\bar{\mu}_{5}(P)$ & $\bar{\mu}_{6}(P)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\mathbb{0}$ & $\imposs$ & $\imposs$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{0}$ & $\necess$ & $\left[\frac{1}{3},1\right]$ & $\left[\frac{2}{3},\frac{2}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{1}$ & $\imposs$ & $\left[0,\frac{1}{3}\right]$ & $\left[\frac{1}{3},\frac{1}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}$ & $\necess$ & $\necess$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
By interpreting the states ~$\proj{0}$ and $\proj{1}$ with the
head and the tail of a coin again, $\bar{\mu}_{4}$ represents tossing
a double-headed coin which can be predicted with certain. When $\delta$
is increased to $\frac{1}{3}$, the classical IVPMs~$\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ defined in Table~\ref{tab:classical-IVPMs}
are clearly $\frac{1}{3}$-deterministic, and they have more indeterminate
behavior than $\bar{\mu}_{4}$. However, the most indeterminate classical
IVPM~$\bar{\mu}_{2}$ defined in Eq.~(\ref{eq:classical-no-information-IVPM})
is still not $\frac{1}{3}$-deterministic because $\bar{\mu}_{2}(\proj{0})=\bar{\mu}_{2}(\proj{1})=\unknown$,
but neither $\unknown\subseteq\left[0,\frac{1}{3}\right]$ nor $\unknown\subseteq\left[1-\frac{1}{3},1\right]$.
Finally, when $\delta=1$, because every interval is a subset of $\unknown$,
every QIVPM is $1$-deterministic. In particular, the informativeness
QIVPM~$\bar{\mu}_{1}$ defined in Eq.~(\ref{eq:no-information-IVPM})
and the Born QIVPM~$\bar{\mu}_{3}$ defined in Eq.~(\ref{eq:Born-QIVPM})
are both $1$-deterministic.

@@

\noindent The observables are constructed using the Pauli matrices
$\left\{ \mathbb{1},\sigma_{x},\sigma_{y},\sigma_{z}\right\}$ whose
eigenvalues are all either 1 or $-1$. They are arranged such that in
each row and column, \emph{except the column $j=2$}, every observable
is the product of the other two. In the $j=2$ column, we have instead
that
$\left(\sigma_{z}\otimes\sigma_{z}\right)\left(\sigma_{x}\otimes\sigma_{x}\right)=-\sigma_{y}\otimes\sigma_{y}$. Now
assume a 0-deterministic QIVPM ${\bar{\mu}}$; the expectation values
of the observables in each row relative to this 0-deterministic QIVPM
are fully determinate and must lie in either the interval $[1,1]$ or
the interval $[-1,-1]$ depending on which eigenvalue is the one whose
associated projector is certain. Since the product of any two
observables in a row is equal to the third, there must be an even
number of occurrences of the interval $[-1,-1]$ in each row and hence
in the entire table. However, looking at the expectation values of the
observables in each column, there must be an even number of
occurrences of the interval $[-1,-1]$ in the first two columns and an
odd number in the $j=2$ column and hence in the entire table. The
contradiction implies the non-existence of the assumed 0-deterministic
QIVPM. We therefore prove a stronger statement of contextuality that
includes the effects of finite-precision.

@@

It remains to verify that $\bar{\mu}^{\textrm{D}}$ is indeed a
0-deterministic QIVPM by checking that for orthogonal
operators~$P_{0}$ and $P_{1}$, we have that
$\bar{\mu}^{\textrm{D}}\left(P_{0}\right)=\bar{\mu}^{\textrm{D}}\left(P_{1}\right)=\imposs$
implies
$\bar{\mu}^{\textrm{D}}\left(P_{0}+P_{1}\right)=\imposs$. Indeed when
$\bar{\mu}^{\textrm{D}}\left(P_{0}\right)=\bar{\mu}^{\textrm{D}}\left(P_{1}\right)=\imposs$,
we have both $\bar{\mu}\left(P_{0}\right)$ and
$\bar{\mu}\left(P_{1}\right)\subseteq\left[0,\delta\right]$ which
implies
$\bar{\mu}\left(P_{0}+P_{1}\right)\subseteq\left[0,2\delta\right]$.
When $\delta<\frac{1}{3}$, i.e., $2\delta<1-\delta$, it follows that
$\bar{\mu}\left(P_{0}+P_{1}\right)$ cannot be a subset of
$\left[1-\delta,1\right]$, but $\bar{\mu}\left(P_{0}+P_{1}\right)$ is
a subset of either $\left[0,\delta\right]$ or
$\left[1-\delta,1\right]$.  Hence, $\bar{\mu}\left(P_{0}+P_{1}\right)$
must be a subset of $\left[0,\delta\right]$, which implies
$\bar{\mu}^{\textrm{D}}\left(P_{0}+P_{1}\right)=\imposs$. Thus the
theorem holds when $\delta<\frac{1}{3}$.

@@

>>>>

In the remainder of this section, we will show that our definition of
QIVPMs is consistent with classical IVPMs in the sense that a
restriction of QIVPMs to commuting events recovers the definition of
classical IVPMs. To formalize this connection between QIVPMs and
classical IVPMs, we first note that in contrast to the case with
conventional quantum probability spaces, there is generally no unique
state~$\rho$ consistent with a QIVPM. We say that a QIVPM $\bar{\mu}$
is consistent with a state $\rho$ and projector~$P$ if the interval
$\bar{\mu}(P)$ contains the exact probability calculated by the Born
rule, i.e, if:
\begin{equation}
\mu_{\rho}^{\mathrm{B}}\left(P\right)\in\bar{\mu}\left(P\right)
\end{equation}
where $\mu_{\rho}^{\mathrm{B}}\left(P\right)=\Tr\left(\rho P\right)$
is the probability that the event~$P$ is determined by the Born
rule~\citep{Born1983,peres1995quantum,544199}.

@@

This theorem follows directly from our
correspondence~\cite{HOSTunpublished} and Shapley's
theorem~\cite{Shapley1971,GilboaSchmeidler1994,Grabisch2016}, which
proves that a classical IVPM always contains at least one state $\rho$
that is consistent with every event in $\events_C$. In contrast, as we
illustrate in Sec.~\ref{sec:Gleason}, the set of states
$\coreBorn\left(\bar{\mu},\events'\right)$ that is consistent with
some arbitrary (not necessarily commuting) events $\events'$ and a
particular QIVPM $\bar{\mu}$ may be \emph{empty}.

@@

\noindent Intuitively the expectation value of an observable relative
to a QIVPM $\bar{\mu}$ lies between two possible outcomes, which
themselves lie between the minimum and maximum bounds of the probability
intervals associated with each state~$\rho$ that is consistent with
$\bar{\mu}$ on every projector in the spectral decomposition of the
observable. If $\bar{\mu}$ is a conventional (Born) probability measure
induced by a state $\rho$, then the Born rule probability induced
by every state in $\coreBorn\left(\bar{\mu},\events'\right)$ will
be $\mu_{\rho}^{\mathrm{B}}$ and the interval collapses to a point,
thus reducing the definition to that of Eq.~(\ref{eq:quantum-expectation})~\citep{HOSTunpublished}.
For example, we can extend one of the fragments of valid probability
measures~$\mu_{1}$ defined in Table~\ref{tab:quantum-probability-measure}
to a QIVPM~$\bar{\mu}_{3}$:
\begin{equation}
\begin{aligned}\bar{\mu}_{3}(\proj{0}) & =\left[\tfrac{1}{2},\tfrac{1}{2}\right] & \bar{\mu}_{3}(\proj{2}) & =\left[0,0\right]\\
\bar{\mu}_{3}(\proj{1}) & =\left[\tfrac{1}{2},\tfrac{1}{2}\right] &  & \ldots\textrm{ etc.}
\end{aligned}
\label{eq:Born-QIVPM}
\end{equation}
For the same observable~$\mathbf{O}=\proj{0}+2\,\proj{1}+3\,\proj{2}$,
the only state~$\rho$ consistent with $\bar{\mu}_{3}$ on $\proj{0}$,
$\proj{1}$, and $\proj{2}$ must be $\proj{\ps}$, and $\mu_{1}=\mu_{\proj{\ps}}^{\mathrm{B}}$.
Therefore, the expectation value~$\expval{\mathbf{O}}_{\bar{\mu}_{3}}=\left[\expval{\mathbf{O}}_{\mu_{1}},\expval{\mathbf{O}}_{\mu_{1}}\right]=\left[1.5,1.5\right]$
which is an interval, but not between $0$ and $1$ because it is
not a probability of an event. We also note that, when restricted
to commuting projectors, the definition above is consistent with the
classical notion of the \emph{Choquet integral}~\citep{GilboaSchmeidler1994,Grabisch2016}
which is used to calculate the expectation value of random variables
as a weighted average~\citep{HOSTunpublished}.

@@

\noindent The definition puts no restrictions on the choice of intervals,
only on the assignment of intervals to events. When $\delta=0$, every
event must be assigned a probability either in the interval $\imposs$
or in $\necess$, i.e., every event is completely determined with
certainty. As $\delta$ gets larger, the QIVPM allows for more indeterminate
behavior. To illustrate the examples easier, we restrict ourselves
to the classical subspace of commuting events~$\events_{0}=\left\{ \mathbb{0},\proj{0},\proj{1},\mathbb{1}\right\} $.
If we apply the same definition in the classical case, the classical
IVPM~$\bar{\mu}_{4}$ defined in Table~\ref{tab:classical-IVPMs}
is $0$-deterministic because $\imposs\subseteq\left[0,0\right]$
and $\necess\subseteq\left[1-0,1\right]$.
\begin{table}
\noindent \centering{}\caption{\label{tab:classical-IVPMs}Classical IVPMs~$\bar{\mu}_{4}$, $\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ on $\events_{0}$.}
\begin{tabular}{cccc}
\toprule 
\addlinespace
$P$ & $\bar{\mu}_{4}(P)$ & $\bar{\mu}_{5}(P)$ & $\bar{\mu}_{6}(P)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\mathbb{0}$ & $\imposs$ & $\imposs$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{0}$ & $\necess$ & $\left[\frac{1}{3},1\right]$ & $\left[\frac{2}{3},\frac{2}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{1}$ & $\imposs$ & $\left[0,\frac{1}{3}\right]$ & $\left[\frac{1}{3},\frac{1}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}$ & $\necess$ & $\necess$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
By interpreting the states ~$\proj{0}$ and $\proj{1}$ with the
head and the tail of a coin again, $\bar{\mu}_{4}$ represents tossing
a double-headed coin which can be predicted with certain. When $\delta$
is increased to $\frac{1}{3}$, the classical IVPMs~$\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ defined in Table~\ref{tab:classical-IVPMs}
are clearly $\frac{1}{3}$-deterministic, and they have more indeterminate
behavior than $\bar{\mu}_{4}$. However, the most indeterminate classical
IVPM~$\bar{\mu}_{2}$ defined in Eq.~(\ref{eq:classical-no-information-IVPM})
is still not $\frac{1}{3}$-deterministic because $\bar{\mu}_{2}(\proj{0})=\bar{\mu}_{2}(\proj{1})=\unknown$,
but neither $\unknown\subseteq\left[0,\frac{1}{3}\right]$ nor $\unknown\subseteq\left[1-\frac{1}{3},1\right]$.
Finally, when $\delta=1$, because every interval is a subset of $\unknown$,
every QIVPM is $1$-deterministic. In particular, the non-informative
QIVPM~$\bar{\mu}_{1}$ defined in Eq.~(\ref{eq:no-information-IVPM})
and the Born QIVPM~$\bar{\mu}_{3}$ defined in Eq.~(\ref{eq:Born-QIVPM})
are both $1$-deterministic.

@@

\noindent The observables are constructed using the Pauli matrices
$\left\{ \mathbb{1},\sigma_{x},\sigma_{y},\sigma_{z}\right\}$ whose
eigenvalues are all either 1 or $-1$. They are arranged such that in
each row and column, \emph{except the column $j=2$}, every observable
is the product of the other two. In the $j=2$ column, we have instead
that
$\left(\sigma_{z}\otimes\sigma_{z}\right)\left(\sigma_{x}\otimes\sigma_{x}\right)=-\sigma_{y}\otimes\sigma_{y}$. Now
assume a 0-deterministic QIVPM ${\bar{\mu}}$; the expectation values
of the observables in each row relative to this 0-deterministic QIVPM
are fully determinate and must lie in either the interval $[1,1]$ or
the interval $[-1,-1]$ depending on which eigenvalue is the one whose
associated projector is certain. Since the product of any two
observables in a row is equal to the third, there must be an even
number of occurrences of the interval $[-1,-1]$ in each row and hence
in the entire table~\citep{HOSTunpublished}. However, looking at the expectation values of the
observables in each column, there must be an even number of
occurrences of the interval $[-1,-1]$ in the first two columns and an
odd number in the $j=2$ column and hence in the entire table~\citep{HOSTunpublished}. The
contradiction implies the non-existence of the assumed 0-deterministic
QIVPM. We therefore prove a stronger statement of contextuality that
includes the effects of finite-precision.

@@

It remains to verify that $\bar{\mu}^{\textrm{D}}$ is indeed a
0-deterministic QIVPM by checking~\citep{HOSTunpublished} that for orthogonal
operators~$P_{0}$ and $P_{1}$, we have that
$\bar{\mu}^{\textrm{D}}\left(P_{0}\right)=\bar{\mu}^{\textrm{D}}\left(P_{1}\right)=\imposs$
implies
$\bar{\mu}^{\textrm{D}}\left(P_{0}+P_{1}\right)=\imposs$. Indeed when
$\bar{\mu}^{\textrm{D}}\left(P_{0}\right)=\bar{\mu}^{\textrm{D}}\left(P_{1}\right)=\imposs$,
we have both $\bar{\mu}\left(P_{0}\right)$ and
$\bar{\mu}\left(P_{1}\right)\subseteq\left[0,\delta\right]$ which
implies
$\bar{\mu}\left(P_{0}+P_{1}\right)\subseteq\left[0,2\delta\right]$.
When $\delta<\frac{1}{3}$, i.e., $2\delta<1-\delta$, it follows that
$\bar{\mu}\left(P_{0}+P_{1}\right)$ cannot be a subset of
$\left[1-\delta,1\right]$, but $\bar{\mu}\left(P_{0}+P_{1}\right)$ is
a subset of either $\left[0,\delta\right]$ or
$\left[1-\delta,1\right]$.  Hence, $\bar{\mu}\left(P_{0}+P_{1}\right)$
must be a subset of $\left[0,\delta\right]$, which implies
$\bar{\mu}^{\textrm{D}}\left(P_{0}+P_{1}\right)=\imposs$. Thus the
theorem holds when $\delta<\frac{1}{3}$.

@@

\yutsung{About contextuality or non-contextuality, summarize the
existence and non-existence of $\delta$-deterministic QIVPMs in Fig.~\ref{fig:contextuality-non-contextuality}.}
\begin{figure*}
\[
\xymatrix{ & \delta=0 &  & \delta=\frac{1}{3} & \delta=1\\
\textrm{Classical IVPMs} & \ar[rrr] &  &  & \,\\
\textrm{Instance} & \bar{\mu}_{4} &  & \bar{\mu}_{5},\bar{\mu}_{6} & \bar{\mu}_{2},\textrm{ every classical IVPM}\\
\textrm{Interpertation} & \textrm{D Classical Theory} &  & \textrm{P Classical Theory?} & \textrm{P Classical Theory?}\\
\textrm{QIVPMs} & \ar@2{|-|}[rr]_{\textrm{NE by Thm. \ref{cor:Kochen-Specker-IVPM}}}\ar[rrr] &  &  & \,\\
\textrm{Instance} & \textrm{NE by Thm. }\ref{thm:Kochen-Specker} &  & \iota_{\infty}\circ\mu_{\mathbb{1}/3}^{\mathrm{B}} & \bar{\mu}_{1},\bar{\mu}_{3}\textrm{, every QIVPM}\\
\textrm{Interpertation} & \textrm{NEDNCHVT}\ar@{}[rr]|-{\textrm{NEPNCHVT?}} &  & \textrm{EPNCHVT?} & \textrm{EPNCHVT?}
}
\]
\caption{\label{fig:contextuality-non-contextuality}The figure summarize the
existence and non-existence of $\delta$-deterministic classical IVPMs
and QIVPMs. In the figure, ``D'' denotes ``deterministic,'' P
denotes ``probabilitic,'' ``(N)E'' denotes ``(Non-)Exist,''
``(N)C'' denotes ``(Non-)Contextuality,'' and HVT denotes ``Hidden-Variable
Theory.''}
\end{figure*}
